\def\topic{SetsLectures}
\def\format{print}

\documentclass[10pt]{memoir}
\chapterstyle{ell}
\input{packages}

\input{meta}

\input{defs.tex}

%\input{style.tex}

\newlength\drop
\newcommand*{\mytitle}{
	\begingroup%
		\drop=0.1\textheight
		\fboxsep 0.5\baselineskip
		\sffamily
		{\centering
			{\Large \Course}\par
			\vspace{0.3\drop}
			{\Large Lecture Notes}\par
			\vspace{0.3\drop}
			{\Large on}\par
			\vspace{0.3\drop}
			{\Large \Topic}\par
			\vspace{0.9\drop}
			{\large \Author}\par
			\vspace{0.3\drop}
			{\Date}\par
			\vspace{1\drop}
		}
		\vfill
		\begin{overview}\Overview\end{overview}\par
		\vfill
	{\small Version: \Version\ (\today) Format: \format}
	\endgroup
}

%\newcommand{\setTopic}[1]{\def\Topic{#1}}
%\newcommand{\Date}{October 2014}
%\newcommand{\Overview}{We will study sets and functions}

%\setTopic{Sets and Functions}

\begin{document}	
\pagestyle{empty} 
\mytitle
\clearpage
\pagestyle{simple}
\tableofcontents
\clearpage

\input{\topic.tex}

%\newwrite\metafile
%\immediate\openout\metafile=meta.tex
%\immediate\write\metafile{\noexpand\providecommand{\noexpand\Author}{M. A. M.}}
%\closeout\metafile

\end{document}

% \part{Introduction}

% Recurring ideas in this course.
% \begin{itemize}
% \item Natural numbers. The natural numbers (counting numbers), without
%   question, constitute the fundamental structure in mathematics.
%   Nearly everything else depends on understanding how natural numbers
%   behave. And yet, you have known how to count from a very young
%   age. So it turns out that a great deal of very sophisticated
%   mathematics depends on us going back to basics, and simply looking
%   carefully at some ideas we already know.
% \item Universal constructions. It turns out that a lot of the ideas in
%   mathematics exhibit a general pattern of being ``universal'' for a
%   particular problem. A simple example is the \emph{floor} of a real
%   number $x$. This is the integer (written $\lfloor x\rfloor$) that is
%   as large as possible and still less than or equal to $x$. So the
%   floor of $5.3$ is $5$ and the floor of $-1.2$ is $-2$.  The floor is
%   universal for the problem of finding an integer $m$ so that $m\leq
%   x$. There are infinitely many integers satisfying this inequality,
%   but $\lfloor x\rfloor$ is the only one we need to know because for
%   an integer, $m\leq x$ is exactly the same as $m\leq \lfloor
%   x\rfloor$.
 
%   Many other constructions that have nothing obvious to do with
%   numbers are similar. A \emph{universal construction} solves a
%   particular problem in such a way that all other solutions are
%   related to the universal construction.
% \item Duality. Many ideas is mathematics come in pairs: minimum and
%   maximum, great common divisor and least common multiple, ``and'' and
%   ``or'', union and intersection. All of these are examples of
%   \emph{duals} concepts. Dual concepts are helpful because a lot of
%   what we can learn about one of a dual pair is automatically true of
%   the other.
% \item Concrete mathematics. We do not shy away from concrete
%   calculations. This will be especially so in the last part of the
%   course.
% \end{itemize}

%\part{Numbers}
% 2
%\input{NaturalNumbers.tex}

% 1
%\input{ListsStructures.tex}
%\input{OtherNumbers.tex}

% 1
%\input{PythonPrimer.tex}
%\input{SumAndProduct.tex}

% 1
%\input{Numerals.tex}

% 2
%\input{OrderAndDivisibility.tex}
%\input{MinMaxGcdLcm.tex}

%\input{SetsAndFunctions.tex}

%\end{document}

\chapter{Integers, Rationals and Reals}

\section{The Integers}

Integers solve an accounting problem that natural numbers can not: how
to deal with assets \emph{and liabilities}.  They allow us to account
for Bob having $10$ dollars in his pocket and at the same time owing
 $15$. At that moment, Bob has, in effect, $-5$ dollars. But $-5$ is
meaningless if all we have are natural numbers. 

Imagine that we have a good understanding of counting, but have never
thought about negative numbers. How could we invent the idea of an
integer.  First, let us describe Bob's situation in more detail by
writing $10\dotminus 15$. That is, Bob has $10$ in assets and owes
$15$ in liabilities. We can not use the standard minus sign because we
are not describing an \emph{operation} (subtraction) that produces
$-5$. Instead we are describing a \emph{situation} (a balance) involving two
natural numbers: an asset and a liability.  Now suppose another guy,
Weaver, has $24$ dollars in his pocket, but he owes $29$.  So we
describe Weaver's situation as $24\dotminus 29$. Evidently, Bob and
Weaver are in the same situation as far as their \emph{balance} is
concerned as they both owe $5$ more than they have. We invent integers 
to deal with exactly this abstract accounting of ``balance''. 

\begin{axiom}
  \emph{Integers} are objects satisfying the following three conditions:
  \begin{itemize}
  \item For any two natural numbers, $m$ and $n$, there is an integer
    that we may call the \emph{difference of $m$ and $n$}. For the
    purposes of this discussion, we write this as $m\dotminus n$.
  \item There are no other integers. That is, every integer equals
    $m\dotminus n$ for some natural numbers $m$ and $n$.
  \item Equality of integers is determined by the condition:
    $m\dotminus n = p\dotminus q$ if and only if $m+q = p+n$.
  \item Every natural number is an integer. In particular, $m = m\dotminus 0$.
  \end{itemize}
\end{axiom}


In this axiom, we include a criterion for when two abstract objects are equal.
This raises a question. Namely, is it even possible to have objects for which
equality matches the criterion. At the very least, the criterion for equality
($m\dotminus n = p \dotminus q$ if and only if $m+q = p+n$) should meet three
minimal ``sanity conditions'':
\begin{itemize}
\item $i=i$ should be true for every integer.
\item $i=j$ should imply that $j=i$.
\item $i=j$ and $j=k$ should imply $i=k$
\end{itemize}

Suppose $i = m\dotminus n$ for some natural numbers $m$ and $n$. So 
$i = i$ means the same thing as $m + n = m+n$. This is obviously
true. Suppose $j = p\dotminus q$ for some natural numbers $p$ and $q$.
So $i = j$ means $m+q = p+n$ and $j=i$ means $p+n = m+q$.  Clearly, if
$i=j$ then $j=i$. Finally, suppose $k=r\dotminus s$ for some natural
numbers $r$ and $s$. Hence, $i=j$ and $j=k$ means $m+q = p + n$ and
$p+s = r+q$.  Addition yields $(m+q)+(p+s) = (p+n)+(r+q)$. Rearranging
this by commutativity and associativity, $(m+s)+(p+q) = (r+n)+(p+q)$.
Since addition of natural numbers is cancellative, $m+s = r+n$, which
shows that $i=k$.

So far, it looks like we must write something like $m\dotminus n$
whenever we want to talk about an integer.  Surely, that is more cumbersome 
than necessary, since most folks are perfectly happy writing the
integer $-5$ instead of, say, $10\dotminus 15$. The notation $m\dotminus n$
is intended just for our temporary use while we sort out just what 
an integer is. For the rest of the course, we would prefer to use
standard notation, which still needs some justification.
In our balance notation, the integer positive $5$ can be written as $5\dotminus 0$,
while negative $5$ can be written as $0\dotminus 5$. We can show that
this is a completely general fact. 

\begin{lemma}
  For each integer $i$, there is a unique natural number $a$
  for which $i = a\dotminus 0$ or $i = 0\dotminus a$.
\end{lemma}
\begin{proof}
  Every integer $i$ equals $m\dotminus n$ for some $m$ and $n$, according
  to our axioms.  By induction on $m$, we will show that for every
  $n$, it is true that $m\dotminus n = 0\dotminus a$ or $m\dotminus n
  = a\dotminus 0$ for some suitably chosen natural number $a$.
  \begin{description}
  \item[Basis] For $m=0$, the result is trivial as we can take $a=n$.
  \item[Inductive hypothesis] Suppose that for some fixed $k$ the
    following is true: for all $n$, there is a natural number $a$
    such that $k\dotminus n = 0\dotminus
    a$ or $k\dotminus n = a\dotminus 0$.
  \item[Inductive Step] Our task is to prove that for all $n$, $k^\nxt
    \dotminus n = 0 \dotminus a$ or $k^\nxt \dotminus n = a \dotminus
    0$ for some $a$. There are two cases to consider: either $n = 0$
    or $n = p^\nxt$ for some natural number $p$.  In the first case,
    the result is trivial as we can take $a = k^\nxt$.  In the second
    case, $k^\nxt \dotminus p^\nxt = k\dotminus p$ (because $k^\nxt +
    p = k + p^\nxt$).  So by the inductive hypothesis, 
    $k^\nxt\dotminus p^\nxt = 0\dotminus a$ or $k^\nxt\dotminus p^\nxt =
    a\dotminus 0$ for some $a$.
  \end{description}

  Finally, we need to show that the choice of $a$ is not
  arbitrary. Suppose $i$ is an integer, and that $i = 0\dotminus a_0$
  and $i = 0\dotminus a_1$. This means that $0+a_0 = n$ and $0 + a_1 =
  n$.  So $a_0 = a_1$. Suppose $i =
  a_0 \dotminus 0$ and $i=a_1\dotminus 0$. Then similar argument shows
  that $a_0=a_1$. Finally, suppose $i = a_0\dotminus 0$ and $i =
  0\setminus a_1$.  This means that $a_0 + a_1 = 0 + 0$. So 
  $a_0 = a_1 = 0$ (remember that $a_0$ and $a_1$ are natural numbers).
\end{proof}

This result justifies our abandonment of the tedious balance notation
$m\dotminus n$ in favor of just two forms: $a \dotminus 0$ and $0\dotminus a$.
Even better, we can write ${}^+a$ as an abbreviation for $a\dotminus 0$ and ${}^-a$ as
an abbreviation for $0\setminus a$.
A positive integer gets the plus sign in front. Of course,
that is not really since we have also assumed that $m = m\dotminus 0$. 
So we will only write ${}^+a$ to emphasize that we are considering $a$
as an integer, not only as a natural number.

The unique natural number $a$ determined by an integer $i$ is 
called the \emph{absolute value} of $i$, and is written as $|i|$.  For
any $m\neq 0$, ${}^+m \neq {}^-m$. So for any integer other than $0$,
the \emph{sign} ($+$ or $-$) is also determined. In other words, each
integer is exactly one of three types: zero, positive or negative.

Define the \emph{signum} of an integer $i$ to be the integer
${}^\pm0$, ${}^+1$ or ${}^-1$ according to whether $i$ is zero,
positive or negative. The signum is written as $\sgn(i)$.

Addition on integers can be defined in the obvious way. The important
features of the definition are:
\begin{itemize}
\item Addition on integers agrees with addition on natural numbers: ${}^+
  m + {}^+n = {}^+(m+n)$.
\item Addition on integers is associative: $i+(j+k) = (i+j)+k$.
\item Addition on integers is commutative: $i+j = j+i$.
\item The identity for addition is $0$: $i+0 = i$.
\item Positive and negative integers cancel: ${}^+m + {}^-m = 0$.
\end{itemize}

Recalling the idea of balances, addition has an obvious meaning.
Namely, we should add assets and add liabilities. So $(m\dotminus n) + (p\dotminus q)$
should equal $(m+p)\dotminus(n+q)$.

Exercises to show $+$ works.

To define multiplication, we begin by insisting on two requirements:
\begin{itemize}
\item Multiplication on integers agrees with multiplication on natural numbers: ${}^+m\cdot {}^+n = {}^+(mn)$.
\item Multiplication is commutative. 
\item Multiplication of integers distributes over addition.
\end{itemize}

It turns out that there is exactly one operation on integers that satisfies these two requirements.
The following exercises establish this. 

Using the facts we have, we first prove the familiar facts about multiplication: negative times positive is negative,
and so on.
First, fill in the steps to show that $0 = {}^+m\cdot{}^-n + {}^+(mn)$. 
\begin{align*}
  {}^+m\cdot {}^-n &= {}^+m\cdot {}^-n + 0\\
    &= {}^+m\cdot {}^-n + {}^+m\cdot 0\\
    &= {}^+m\cdot {}^-n + {}^+m\cdot ({}^+n + {}^-n)\\
    &= \\
    &= \\
    &= \\
    &= {}^-(mn)
\end{align*}
From this, show that ${}^+m{}^-n = {}^-(mn)$. 
This shows why a positive times a negative is negative.
Because we expect commutativity, a negative times a positive is negative.
Now, we can also prove that a negative times a negative is positive.
\begin{align*}
  0 &= \\
    &= \\
    &= {}^-m\cdot {}^-n + {}^-m\cdot {}^+n\\
    &= {}^-m\cdot {}^-n + {}^-(mn)
\end{align*}

\section{The Rationals}

Rational numbers are analogous to integers except that multiplication is used in place of addition.
That is, for any two integers $i$ and $j$ with $j\neq 0$, there is a rational number $\frac{i}{j}$.
There are no other rational numbers. And $\frac{i}{j} = \frac{k}{\ell}$ holds if and only if 
$i\cdot\ell = k\cdot j$. For an integer $i$, $\frac{i}{{}^+1}$ is the corresponding rational number.
We extend multiplication of integers to rationals by requiring that $\frac{i}{1} \cdot \frac{j}{1} = \frac{ij}{1}$.
We also require reciprocals to behave: $\frac{i}{j}\cdot \frac{j}{i} = \frac{1}{1}$.

From this, we can define multiplication as \[\frac{i}{j} \cdot \frac{k}{\ell} = \frac{ik}{j\ell}\]
As we all know, $\frac{i}{j} + \frac{k}{\ell} \neq \frac{i+k}{j + \ell}$, but why not? 
After all, that is a very simple, easy to remember rule. Why does that operation not 
match up with what we mean by ``addition''? 

\begin{exercise}
In a paragraph or so (you may include displayed equations and equality chains), 
explain why the rule $\frac{i}{j} + \frac{k}{\ell} = \frac{i\ell + kj}{j\ell}$ makes more
sense than the rule $\frac{i}{j} + \frac{k}{\ell} = \frac{i + k}{j + \ell}$. 
\end{exercise}

We do not go into detail confirming that arithmetic operations work as intended for rational numbers because, for the most part,
the confirmations are similar to what we needed for integers.

From now on, we will use the familiar laws of arithmetic without further comment. The point is that rational numbers
are built from natural numbers in a way that ensures they behave well for the four basic arithmetic operations. 


\section{Floor and Ceiling}

There is a trivial operation that takes 
 an integer $i$ and simply treats it as a real. To make this 
clearer, for this discussion let us give this a name: for now, we write $r(i)$ when we want to think of 
an integer $i$ as a real number. Of course, $i$ and $r(i)$ are the same number.
It is only a matter of emphasizing it as an element of $\ZZ$
or of $\RR$.

Evidently, $\fromto r\ZZ\RR$ is a one-to-one, but not onto function. As we know, there are onto, but not one-to-one functions
from $\fromto f\RR\ZZ$ that act as left inverses of $r$: $f(r(n)) = n$ for all integers $n$. But because both $\RR$ and $\ZZ$
have a natural order $\leq$, we can actually pick out a natural left inverse. Namely, for any real number $x$,
let the \emph{floor} of $x$ be the largest integer that is less than or equal to $x$. We typically write the floor of $x$
as $\lfloor x\rfloor$. So for any integer $n$ and any real number $x$, $r(n) \leq x$ is a sensible comparison to make because
$r(n)$ and $x$ are both real numbers. Also $n\leq \lfloor x\rfloor$ is a sensible comparison to make because
$n$ and $\lfloor x\rfloor$ are both integers. Our definition of $\lfloor\ \rfloor$ means that
\[ r(n)\leq x \qquad \mbox{if and only if}\qquad n\leq \lfloor x\rfloor.\]
For example, $\lfloor 5.2\rfloor = 5$ as $5$ is the largest integer that is less than or equal to $5.2$. 
It is tempting to think about $\lfloor x\rfloor$ simply removes the ``decimal'' part of $x$, but this is wrong for 
negative numbers: $\lfloor -5.2\rfloor = -6$ because $-6$ is less than or equal to $-5.2$, but $-5$ is not.

[Put a graphic here.]

To avoid clutter, let us agree to use the variables $m$, $n$ and $p$ to denote integers and $x$, $y$ and $z$ to denote reals.
Also, we can suppress the inclusion function $r$. So rather than writing $x+r(m)$ we just write $x+m$. 

First, it should be clear that $\lfloor x + m\rfloor = \lfloor x\rfloor + m$. We can prove this quite easily:
$\lfloor x\rfloor + m \leq n$ if and only if $x+m \leq n$ if and only if $\lfloor x+m\rfloor \leq n$. 
Also $\lfloor x\rfloor \leq x < \lfloor x\rfloor + 1$. Because $\lfloor x\rfloor \leq \lfloor x\rfloor$,
$\lfloor x\rfloor \leq x$. And because $x+1\nleq x$, $\lfloor x\rfloor + 1 = \lfloor x+1\rfloor \nleq x$.
Since real numbers are totally ordered, $x< \lfloor x\rfloor + 1$.

Writing $\lceil x\rceil$ for $-\lfloor -x\rfloor$. We note that $-x\leq y$ if and only if $-y \leq x$. So
$m\leq \lceil x\rceil$ if and only if $\lfloor -x \rfloor \leq -m$ if and only if $-x\leq -m$ if and only if
$m\leq x$.

Here are some basic facts that are useful for floor and ceiling:

\begin{itemize}
\item $\lfloor\lfloor x\rfloor\rfloor = \lfloor x\rfloor = \lceil\lfloor x \rfloor\rceil$
\item If $x$ and $y$ are positive, then 
\[\lfloor x\rfloor\cdot \lfloor y\rfloor 
    \leq \lfloor x\cdot y\rfloor
    <    \lfloor x\rfloor \cdot \lfloor y\rfloor + \lfloor x\rfloor + \lfloor y\rfloor + 1\]
\end{itemize}

First, $\lfloor x\rfloor \cdot \lfloor y\rfloor \leq x\cdot y$. So $\lfloor x\rfloor \cdot \lfloor y\rfloor \leq \lfloor x\cdot y\rfloor$.
Also, $\lfloor x\cdot y\rfloor \leq x\cdot y< (\lfloor x\rfloor + 1)(\lfloor y\rfloor + 1)$.

$\lfloor x^2\rfloor \leq x^2 < \lfloor x\rfloor^2 + \lfloor x\rfloor$ because $x^2 = \lfloor x\rfloor^2 + 2x\{x\} + \{x\}^2$. 
 

\subsection{Rounding in General}

Most of us were taught to \emph{round} a real number by taking the ``nearest'' integer. So $1.2$ rounds to $1$ and $1.6$ rounds to $2$.
In cases like $1.5$, we round up to $2$, but the reason may seem a little arbitrary. Actually, the thinking is that when we write $1.5$
we probably really mean that we are simply writing the first $2$ digits of a longer decimal expansion. So $1.5$ might really be $1.500001$.
In that case, it should definitely round up. Anyway, this is the agreed-upon convention.

To round a real number to the nearest integer, we simply take $\lfloor x+1/2\rfloor$. Rewriting, this is $\lfloor (2x+1)/2\rfloor$.
The result is that a number with fractional part less than $1/2$ rounds down and a number with fractional part greater than or equal to $1/2$ rounds up.

Suppose we wish to truncate a real number to a multiple of $10$ instead of to an integer. So $123.45$ would truncate to $120$.
The procedure is simple:  first, divide by $10$, then round, then multiply by $10$. So the formula $10\lfloor x/10\rfloor$ does the job.
In fact, for any positive real $c$, $c\lfloor x/c\rfloor$ truncates $x$ to a multiple of $c$.  

\begin{exercise}
  Calculate $c\lfloor x/c\rfloor$ where $c = 1/4$ for the following values of $x$: $3.1$, $5.25$, $-1.1$, $109230$. Show every step.
\end{exercise}

Suppose, for some reason, we wanted to ``round'' real numbers to that anything with fractional part below $1/3$ rounds down and
anything with fractional part greater than or equal to $1/3$ rounds up. So $2.33$ would round down to $2$, whereas $2.34$ would round up to $3$. Let us call this ``rounding at $1/3$''. Notice that the conventional rounding (rounding at $1/2$)
is defined by $\lfloor (2x+1)/2\rfloor$, which involves two constants: $2$ appears in the denominator
and as a factor next to $x$ in the numerator; $1$ appears in the numerator. So we might consider all formulas of the form $\lfloor (dx + k)/d\rfloor$,
where $d$ is a positive natural number and $k$ is a natural number so that $0\leq k < d$.

\chapter{Lists}

\chapter{Sets and functions}

\chapter{Some Number Theory}

\chapter{Some Combinatorics}

\chapter{Some Sums and Differences}

\chapter{Polynomials}

\chapter{Recurrences}


\part{Everyday Recipes}

\part{Formal Dining}

%\include{codes}
%\include{Intro}
%\include{NumbersNotes}
%\include{FiniteSets}
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
